\documentclass{article}

\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}

\newtheorem{Thm}{Theorem}
\newtheorem{Conj}{Conjecture}
\newtheorem{Cor}{Corollary}
\newtheorem{Lem}{Lemma}
\newtheorem{Prop}{Proposition}
\newtheorem{Ques}{Question}
\newtheorem{Goal}{Goal}
\newtheorem{Claim}{Claim}
\newtheorem{Def}{Definition}
\newtheorem{Rem}{Remark}
\newtheorem{Exa}{Example}
\newcommand{\A}{\mathbb{A}}
\newcommand{\If}{\mbox{if}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\q}{\mathbb{Q}}
\newcommand{\qss}{\ensuremath{{\bf QI}_m}}
\newcommand{\QI}{\ensuremath{ Q_T^{~k,m}}}
\newcommand{\Qdown}{\ensuremath{ Q_T^{~n-i+k,m-1}}}
\newcommand{\tn}{\mathbb{T}_n}
\newcommand{\C}{\mathbb{C}}
\newcommand{\g}{\mathfrak{g}}
\newcommand{\PP}{\mathfrak{p}}
\newcommand{\etal}{{\it et~al.\ }}
\newcommand{\ie} {{\it i.e.,\ }}
\newcommand{\eg} {{\it e.g.,\ }}
\newcommand{\cf}{{\it cf.,\ }}
\newcommand{\eqdef}{\mathbin{\stackrel{\rm def}{=}}}
\newcommand{\p}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\n}{\mathbb{N}}
\newcommand{\z}{\mathbb{Z}}
\newcommand{\f}{\mathbb{F}}
\newcommand{\sgn}{\rm{sgn}}
\newcommand{\hC}{\widehat{C}}
\newcommand{\hR}{\widehat{R}}
\newcommand{\al}{\alpha_{i,j_a}}

\begin{document}
\title{Schoof's Algorithm for Counting Points on $E(\f_q)$}
\author{Gregg Musiker}
\date{December 7, 2005} \maketitle


\section{Introduction}

In this write-up we discuss the problem of counting points on an elliptic curve over a finite
field.  Here, an elliptic curve $E$ is the zero locus of an algebraic equation of a special form.
Over a field of characteristic $\not = 2, 3$, this equation can be written as \begin{eqnarray}
\label{WeEq} y^2 = x^3 + Ax+B.\end{eqnarray} This is commonly known as {\bf Weierstra\ss ~form}. To
be rigorous, we would have to introduce projective coordinates to define the entire zero locus.
However, in the case of an elliptic curve, the zero locus includes all the points $(x,y)$
satisfying equation (\ref{WeEq}), these are known as affine points, plus \emph{exactly one} additional
point $P_\infty$ ``at infinity".  While there is also an extensive theory of elliptic curves over
$\q$ and $\C$, in this presentation, we will focus on elliptic curves $E(\f_q)$ over a finite field
$\f_q$, where $q=p^k$ and $p$ is a prime greater than $3$.

One of the important properties of elliptic curves is the existence
of a group law $$\oplus : E \times E \rightarrow E.$$ Given $E$ as
above, the addition on the curve is given as follows: For $P_1 =
(x_1,y_1)$ and  $P_2 = (x_2,y_2)$, $P_1 \oplus P_2 = P_3 =
(x_3,y_3)$ where

1) If $x_1 \not = x_2$ then $$x_3 = m^2 - x_1 - x_2 \mathrm{~~and~~}
y_3 = m(x_1-x_3)-y_1 \mathrm{~~with~~} m= {y_2-y_1 \over x_2-x_1}.$$

2) If $x_1 = x_2$ but ($y_1\not = y_2$, or $y_1=0=y_2$)  then $P_3 = P_\infty$. \\

3) If $P_1 = P_2$ and $y_1 \not = 0$, then
$$x_3 = m^2 - 2x_1 \mathrm{~~and~~} y_3 = m(x_1-x_3)-y_1 \mathrm{~~with~~} m= {3x_1^2+A \over 2y_1}.$$

4) $P_\infty$ acts as the identity element in this addition.

\noindent Among other things, this allows elliptic curves to be used
in cryptography, usually taking advantage of the difficulty of the
discrete logarithm problem for group $E(\f_q)$. Because of this
application, an important computation for cryptographic purposes is
the cardinality of $E(\f_q)$, which also happens to be the order of
the group used for encryption.  If the order of the group has small
primes dividing it, then it leads to a weaker encryption scheme.
Thus, if one is going to use elliptic curves for cryptography, it is
important to have an efficient method to test the strength of the encryption scheme for a
given curve $E$ over $\f_q$.

While a few methods exist for computing the size of $|E(\f_q)|$ (as
described in \cite{Harris}), e.g. directly counting after running
through possible values of $x$ and $y$ in $\f_q$, applying the
formula $$|E(\f_q)| = q + 1 + \sum_{\alpha \in \f_q} \bigg({x^3+Ax+B
\over q}\bigg), $$ Shank's Baby Step-Giant Step
Method; the fastest one for extremely large primes is
a variant of Schoof's algorithm.

Here I will first describe Schoof's original algorithm, as presented
in \cite{Schoof1} and summarized in \cite{Wash}.  After this
warm-up, we will describe an improvement of Elkies which leads to a
more efficient algorithm.  There is in fact an additional
improvement from Atkin to make Schoof-Elkies-Atkin (SEA) one of the
fastest algorithm for counting the number of points on $E$ over a
large prime field, but its description will be outside the scope of
this paper.  (This full algorithm has been implemented in the
computer algebra systems Magma, PARI, and SAGE.  Please see
\cite{PariImp} for details on the PARI implementation.  In SAGE, you
can use the command $E.sea(p)$.)

To begin, we note that the size of $E(\f_q)$ is very close to $q+1$.  This is not completely surprising
since half of the elements of $\f_q^\times$ are squares, and the equation $y^2 = \alpha$ has two solutions
if $\alpha$ is a square, one solution if $\alpha=0$, and zero solutions if $\alpha$ is not a square
in $\f_q^\times$.  Thus $y^2 = x$ would have exactly $q$ solutions of form $(x_0,y_0)$ plus the point at
infinity. Unfortunately, it is not always true that quantity $x^3+Ax+B$, $x \in \f_q$, is a square half
the time, but we at least expect it to be a square \emph{close} to half the time.  By analogous
reasoning, we get that $y^2 = x^3+Ax+B$ will have approximately $q+1$ solutions, including
$P_\infty$.  

We now proceed to make this statement more precise.
\begin{Thm} [Hasse 1934] \label{HasseWeil}
Letting $N_1$ denote $|E(\f_q)|$, we obtain that $$|N_1 - q -1| \leq 2\sqrt{q}.$$
\end{Thm}
\noindent This is a deep theorem, and can actually be shown to be equivalent to the Riemann Hypothesis for
genus one function fields.  We will give a partial proof of this theorem in Section \ref{FrobChar}.  More details are in \cite{Wash}.
 We have now simplified our problem by narrowing down $|E(\f_q)|$ to a finite (albeit large) set of
possibilities.

Rene Schoof's insight was exploiting the fact we know that there is a finite range of
possible values for the cardinality of $E(\f_q)$.  Hence, if we had some way of computing
the size of $E(\f_q)$ modulo $N$ where $N > 4\sqrt{q}$, that would be sufficient for determining $|E(\f_q)|$.  There is not an efficient way to compute
$|E(\f_q)|$ mod $N$ directly for general $N$, however there is an efficient way to compute $|E(\f_q)|$ mod $l$ for $l$ prime.

Hence, we will compute $|E(\f_q)|$ mod $l$ for $l$ in a set $S$ of primes such that $\prod_{l \in S} l = N > 4\sqrt{q}$.  The Chinese Remainder Theorem, 
found described in numerous sources including 
\cite{CRT}, allows us to compute $|E(\f_q)|$ mod $N$ given $|E(\f_q)|$ mod $l$ for all $l \in S$.

We now proceed to describe how to efficiently compute $|E(\f_q)|$ mod $l$ for a prime $l \not = $p.  Note that requiring $p \not\in S$ is no loss since
we can pick a bigger prime to take its place to ensure the product is big enough.

To efficiently compute $|E(\f_q)|$ mod $l$, we will acutally find it easier to compute the values
$$t_l \equiv q + 1 -|E(\f_q)|~(\mathrm{mod~} l).$$
This computation will require a digression into the theory of the Frobenius map and Division
Polynomials so we provide those now.

\section{The Characteristic Equation for the Frobenius Endomorphism} \label{FrobChar}

Given an elliptic curve $E(\f_q)$ we consider $\overline{E}$ to be the curve with same defining equation as $E$, but now allowing points with coordinates
in $\overline{\f_q}$, the algebraic closure of $\f_q$.  Given curve $\overline{E}$, there exists a map, in fact its an endomorphism back onto
$\overline{E}$, defined as
$$\pi : (x,y) \mapsto (x^q,y^q).$$
This map has several useful properties:

$\bullet$ For $P \in \overline{E}$, $\pi(P) \in \overline{E}$, thus it is in fact an endomorphism of $\overline{E}$ as claimed.

$\bullet$ For $P \in \overline{E}$, $\pi(P) = P$ if and only if $P \in E(\f_q)$.  In particular $\pi(P_\infty) = P_\infty$.

$\bullet$ For $P,Q \in \overline{E}$, $\pi(P\oplus Q) = \pi(P) \oplus \pi(Q)$, thus $\pi$ is compatible with the addition on the curve and in fact
we can think of acting by $\pi$ as multiplication in our endomorphism ring since the distributive law holds.

On $E(\f_q)$, the Frobenius map satisfies the characteristic equation $\pi - 1 = 0$.  This begs the question of what kind of characteristic equation
$\pi$ satisfies on all of $\overline{E}$.  In fact in satisfies a quadratic equation.

\begin{Thm} \label{chareqn}
Frobenius map $\pi$ satisfies the characteristic equation \begin{eqnarray} \label{eqnchar} \pi^2 - t\pi + q = 0\end{eqnarray} where $t = q + 1
- |E(\f_q)|$.
\end{Thm}

\noindent The proof of Theorem \ref{chareqn} is outside the scope of this presentation, although please
see \cite{Wash} for details.  Nonetheless, we can use Theorem \ref{chareqn} to give a sketch of
the proof of Theorem \ref{HasseWeil}.

\begin{proof} (Sketch) It can be shown that not only is $x^2-tx+q$ the characteristic equation for
the Frobenius map $\pi$, but furthermore, for all rational numbers $r/s \in \q$,
\begin{eqnarray} \label{rateqn} (r/s)^2-t(r/s) + q \geq 0.\end{eqnarray}  This of course implies that $x^2 - tx + q \geq 0$ for all real
numbers $x \in \R$. One way to show inequality (\ref{rateqn}) is by using $\pi$'s characteristic
equation to prove that for all $r,s\in \z$, the quantity $r^2 - trs + qs^2$ equals deg $(r-s\pi)$,
which is in $\z_{\geq 0}$ by definition of degree. Once we have $x^2 - tx + q \geq 0$ for all real
numbers $x \in \R$, Theorem \ref{HasseWeil} quickly follows by noting that the discriminant cannot
be positive, hence
$$t^2 - 4q \leq 0 \Rightarrow |t| \leq 2\sqrt{q}.$$
\end{proof}

Theorem \ref{chareqn} also immediately implies the following identity on the curve $\overline{E}$:
$$\mathrm{For~all~}P=(x,y) \in \overline{E},~~~~(x^{q^2},y^{q^2}) \oplus q(x,y) = t(x^q,y^q)$$ where scalar multiplication by $t$ (or $q$)
signifies adding a point to itself $t$ (or $q$) times.  We now spend the rest of this section, as
well as the next two, on the problem of determining $t_l$, defined as $t$ mod $l$, for a given
prime $l \not = 2, ~p$. We therefore, refer to $l$ as if it's been fixed, unless otherwise
specified.

If point $(x,y)$ is in the torsion subgroup $\overline{E}[l]$, meaning that $l(x,y) = P_\infty$,
then the point $qP = \overline{q}P$ where $\overline{q}$ signifies $q$ mod $l$, choosing
$\overline{q}$ so that $|\overline{q}| < l/2$. Since $\pi(P_\infty) = P_\infty$, and $r \in \z$
implies $r \cdot\pi(P) = \pi(rP)$ by repeated use of the distributive law, $\pi(P)$ will have the
same order as $P$.  Thus for $(x,y) \in \overline{E}[l]$, we also have $ t(x^q,y^q) =
\overline{t}(x^q,y^q)$ where $\overline{t}$ is $t$ mod $l$. Hence we have reduced our problem to
solving the equation
\begin{eqnarray} \label{Ttosol} (x^{q^2},y^{q^2}) \oplus \overline{q}(x,y) \equiv
\overline{t}(x^q,y^q) \mathrm{~(mod~}l)\end{eqnarray} for $\overline{t}$ (mod $l$).

The idea now is to explicitly compute $(x^{q^2},y^{q^2}) \oplus \overline{q}(x,y)$ as a pair of
rational functions $(x^\prime,y^\prime)$ in terms of $x$ and $y$.  One method for finding $t_l$
would involve plugging in $\overline{t} = 0, 1, 2, \dots, l-1$ and find $\overline{t}$ such that
the pair of rational functions given by $\overline{t}(x^q,y^q)$ are the same as the pair of
rational functions on the left-hand-side, thus determining $t_l$.  However, to do this more
efficiently, we use division polynomials to allow us to compute multiples of a point $P$, and work
with polynomials with bounded degrees instead of rational functions.

\section{Division Polynomials} \label{DivPol}

We define a sequence of polynomials in $\z[x,y,A,B]$ via the following initial conditions and recurrence equations:
\begin{eqnarray*}
\psi_0 &=& 0 \\
\psi_1 &=& 1 \\
\psi_2 &=& 2y \\
\psi_3 &=& 3x^4 + 6Ax^2 + 12Bx-A^2 \\
\psi_4 &=& 4y(x^6 + 5Ax^4 + 20Bx^3 - 5A^2x^2 - 4ABx - 8B^2 - A^3) \\
&\cdots& \\
\psi_{2m+1} &=& \psi_{m+2}\psi_m^3 - \psi_{m-1}\psi_{m+1}^3 \mathrm{~for~ }m \geq 2 \\
\psi_{2m} &=& ({\psi_m \over 2y})\cdot (\psi_{m+2}\psi_{m-1}^2 - \psi_{m-2}\psi_{m+1}^2)  \mathrm{~for~ }m \geq 2
\end{eqnarray*}
The polynomial $\psi_n$ is known as the $n$th Division Polynomial \cite{Lang,Wash}.  These
polynomials turn out to have the remarkable property that all of the finite $n$-torsion points
$(x_0,y_0)$, i.e. elements of $\overline{E}[n] \setminus \{P_\infty\}$, satisfy $\psi_n^2(x_0,y_0)
= 0$.  (Note that it can be shown using a variant of the characteristic equation (\ref{eqnchar}) that the number of finite torsion 
points is exactly $n^2-1$.)

Additionally, we can define the multiple of a point, $r\cdot(x,y)$, as a pair of rational functions
in terms of $x$ and $y$ using the $\psi_n$'s. In particular, we have the following:

\begin{Prop} \label{elladd}
Let $P= (x,y)$ be a point on the elliptic curve $y^2=x^3+Ax+B$ over some field of characteristic $\not = 2$.  Then for any positive integer
$n$, $nP = P \oplus P \oplus P \oplus \dots \oplus P$ is given by

$$nP = \bigg({ \phi_n(x)\over \psi_n^2(x)} , ~{\omega_n(x,y) \over \psi_n^3(x,y)}\bigg) =\bigg(x-{\psi_{n-1}\psi_{n+1}\over \psi_n^2(x)} ,
~{\psi_{2n}(x,y) \over 2\psi_n^4(x)}\bigg).$$

$$-nP = \bigg({\phi_n(x)\over \psi_n^2(x)} , ~-{\omega_n(x,y) \over \psi_n^3(x,y)}\bigg) = \bigg(x-{\psi_{n-1}\psi_{n+1}\over \psi_n^2(x)} ,
~-{\psi_{2n}(x,y) \over 2\psi_n^4(x)}\bigg)$$
where the polynomials $\phi_n$ and $\omega_n$ are defined as
\begin{eqnarray*}
\phi_m &=& x\psi_m^2 - \psi_{m+1}\psi_{m-1} \\
\omega_m &=& {\psi_{m+2}\psi_{m-1}^2 - \psi_{m-2}\psi_{m+1}^2 \over 4y}.
\end{eqnarray*}

\end{Prop}
Note that using the equivalence relation $y^2 \equiv x^3 + Ax+B$ and
the recurrence relations for $\psi_{2m}$ and $\psi_{2m+1}$, we can
inductively prove that $$\psi_n^2, ~{\psi_{2n} \over y},
~\psi_{2n+1},~\mathrm{and~}\phi_n
\mathrm{~are~all~functions~in~terms~of~}x.$$ As a corollary, the
$x$-coordinate of $nP$ is a rational function strictly in terms of
$x$, and the $y$-coordinate has the form $y\cdot\Theta(x)$.
Proposition \ref{elladd} is commonly proved using the theory of the
Weirestra\ss~$\mathfrak{P}$-function, as given in \cite{Lang} and
\cite[Chapter 9]{Wash}.


We can summarize these results as follows: $\psi^2$ is a function in
$x$ alone and has degree $n^2-1$, which equals the number of finite
$n$-torsion points.  The degree of $\psi^2$ is easily verified via
the above recurrence relations.  Furthermore, if $n$ is odd and
$(x_0,y_0) \in \overline{E} \setminus \{P_\infty\}$, then
\begin{eqnarray} \label{oddpsi} \psi_n(x_0) = 0
\mathrm{~if~and~only~if~}(x_0,y_0) \in
\overline{E}[n].\end{eqnarray}

If $n$ is even, $E$ defined by equation $y^2 = (x-\alpha_1)(x-\alpha_2)(x-\alpha_3)$ over
$\overline{\f_q}$, and $(x_0,y_0) \in \overline{E} \setminus \{P_\infty,
~(\alpha_1,0),~(\alpha_2,0),~(\alpha_3,0)\}$, then

\begin{eqnarray} {\psi_n \over y}(x_0) = 0 \mathrm{~if~and~only~if~}(x_0,y_0) \in
\overline{E}[n].\end{eqnarray}

%\begin{Ques}
%Given a curve over a finite field, is a simpler proof known for proving these properties that does
%not require the theory of elliptic curves over $\C$ and the notion that identities over
%characteristic $0$ imply identities over characteristic $p$?  After all, it seems strange that a
%direct proof which only requires the theory of elliptic curves over finite fields does not exist.
%Closest I have found is \cite{Enge}.  This is an unenlightening technical proof however. [FOR DRAFT
% AND OH ONLY UNLESS A REASONABLE QUESTION]
%\end{Ques}

%\begin{Lem}\cite[pg. 78]{Wash}

%\begin{eqnarray*}
%\phi_n(x) &=& x^{n^2} + \mathrm{~lower~degree~terms}\\
%\psi_n^2(x) &=& n^2x^{n^2-1} + \mathrm{~lower~degree~terms}\\
%\mathrm{in ~fact} \\
%\psi_{2n} &=& y(2n x^{2n^2-2} + \mathrm{~lower~degree~terms}\\
%\psi_{2n+1} &=& (2n+1)x^{2n^2+2n} + \mathrm{~lower~degree~terms}
%\end{eqnarray*}
%\end{Lem}


%\begin{Cor}
%The degree of the endomorphism/isogeny of multiplication by $n$ has degree $n^2$.
%\end{Cor}

%This is simply because the maximum of the degrees of $\phi_n(x)$ and $\phi_n^2(x)$, which in fact only depend on $x$, is $n^2$.

%Possibly more clearly,
%the degree of the denominator is $n^2-1$ and this morphism/isogeny is separable, which means it has no multiple roots, and thus we have $n^2-1$ values
%of $\alpha \in \overline{\f_p}$ we can plug in to obtain a zero denominator, i.e. an $x$-coordinate of $\infty$.

%Hence, if we let $P = P_\infty$ or $(\alpha, \beta)$ where $\alpha$ a zero of $\phi_n^2(x)$, we obtain $nP = P_\infty$.  $n^2$ possibilities, thus
%$n^2$ elements in the kernel of this seperable morphism, hence degree $n^2$.

%Note in the case $gcd(n,p) > 1$ the multiplication map is not separable.  The degree is still $n^2$, but the size of the kernel is smaller since
%multiple roots.

\section{The Remainder of the Original Algorithm}

With preliminaries out of the way, we proceed to present Schoof's algorithm for computing $t_l$,
which is defined as $q + 1 - |E(\f_q)|$ (mod $l$), for $l \not = 2,~p$.  We recall our definition
of $\overline{q}$ as a specific integer satisfying $\overline{q} \equiv q$ (mod $l$) and
$|\overline{q}| < l/2$.  First we use division polynomials to rewrite $\overline{q}(x,y)$ as a pair
of rational functions in $x$ and $y$:

$$(x_{\overline{q}}, y_{\overline{q}}) = \overline{q}(x,y) = \bigg(x-{\psi_{\overline{q}-1}\psi_{\overline{q}+1}\over \psi_{\overline{q}}^2(x)} ,
~{\psi_{2\overline{q}}(x,y) \over \psi_{\overline{q}}^4(x)}\bigg).$$

We then encounter an obstacle because the formula for computing
$$(x^{q^2},y^{q^2}) \oplus \overline{q}(x,y),$$ given rational
function representations of each of $(x^{q^2},y^{q^2})$ and
$(x_{\overline{q}},y_{\overline{q}}) = \overline{q}(x,y)$, will
depend on which of three cases, (e.g. are the points distinct, do
they share the same $x$-coordinate), we are in.  There is not an efficient enough way to determine which case we are to justify taking the time to check 
whether or not we happen to fall into the two uncommon special cases.  Thus we might as well assume we happen to be in the case 
$(x^{q^2},y^{q^2}) \not = \pm \overline{q}(x,y)$.  If our guess is correct, 
which it will be most of the time, the proceeding algorithm will find $t_l$ for us.  If our guess is incorrect, our algorithm 
will find that no such $\overline{t}$ that will satisfy (\ref{Ttosol}), hence alerting us that we are in the case
$(x^{q^2},y^{q^2}) = \overline{q}(x,y)$ or  $(x^{q^2},y^{q^2}) = - \overline{q}(x,y)$.  We will prove the above assertions, and give the algorithm for 
these two cases in Section \ref{DecCases}.

Thus, as prescribed, we now assume that
$(x^{q^2},y^{q^2}) \not = \pm \overline{q}(x,y)$ for some $(x,y) \in
\overline{E}[l]\setminus\{P_\infty\}$.  Hence, if we wish to compute
$$(x^\prime, y^\prime) = (x^{q^2},y^{q^2}) \oplus \overline{q}(x,y),$$ we can use the usual addition formula and obtain
$$x^\prime = \bigg({y^{q^2}-y_{\overline{q}} \over x^{q^2}-x_{\overline{q}}}\bigg)^2 - x^{q^2}- x_{\overline{q}}.$$
We get an analogous formula for $y^\prime$, but it is more
computationally efficient to first use the $x$-coordinate to narrow
down the choice of $\overline{t}$ to two possibilities. We must then
determine which square-root of
$$\pm \sqrt{{x^\prime}^3 + Ax^\prime + B}$$ to use.

We recall from Section \ref{DivPol} that the $y$-coordinate of
$\overline{q}(x,y)$, which we denote as $y_{\overline{q}}$, equals
$y\Theta(x)$ and factor
$$(y^{q^2}-y\Theta(x))^2 = y^2(y^{q^2-1}-\Theta(x))^2 =
(x^3+Ax+B)\bigg((x^3+Ax+B)^{q^2-1 \over 2}-\Theta(x)\bigg)^2$$ to
discover that $x^\prime$ is a rational function in variable $x$
alone.

We will soon use the following fact: If $x^\prime = x_{\overline{t}}^q$ for one point $P$ in
$\overline{E}[l]\setminus \{P_\infty\}$, then $\overline{t}$
satisfies
$$\pi^2(P) \ominus \overline{t}\pi(P) \oplus qP = P_\infty.$$ But
since $\overline{t}$ in the characteristic equation is fixed, we
obtain that this choice of $\overline{t}$ must indeed be $t_l$ and
we have proven $x^\prime = x_{\overline{t}}^q$ for all points $P$ in
$\overline{E}[l]\setminus \{P_\infty\}$.

We recall that our goal is to solve (\ref{Ttosol}) for
$\overline{t}$.  Taking $x$-coordinates of both sides, we obtain
that the left-hand-side of (\ref{Ttosol}) is $x^\prime$ and the
right-hand-side is $x_{\overline{t}}^q$.  We use our fact that for
odd prime $l$, point $(x_0,y_0) \in \overline{E}[l]\setminus
\{P_\infty\}$ if and only if $\psi_l(x_0)=0$.  Furthermore, the
roots of $\psi_l$ are simple by an easy counting argument.  In
particular, the degree of $\psi_l^2$ is exactly the number of finite
$l$-torsion points.  

Thus the equality $x_0^\prime =
x_{0_{\overline{t}}}^q$ at all points $(x_0,y_0) \in
\overline{E}[l]\setminus \{P_\infty\}$ is equivalent to the
statement $$\psi_l(x) \bigg | (x^\prime - x_{\overline{t}}^q)$$ as a 
polynomial in $x$, modulo $l$.  In conclusion, to solve (\ref{Ttosol}) for $\overline{t}$, we need only solve
\begin{eqnarray} \label{Eqtosol} x^\prime - x^q_{\overline{t}} \equiv 0 ~(\mathrm{mod~}\psi_l)\end{eqnarray} for
$\overline{t}$.

This computation can be done efficiently for given $\overline{t} \in
\{1,2,\dots,{l-1\over 2}\}$ by computing the power $x_{\overline{t}}^q$ by
{\bf successive squaring}.  (Note that we can stop checking $\overline{t}$ at ${l-1 \over 2}$ since we are only currently determining $t_l$ up 
to additive inverse modulo $l$.)  We create a table of
$$x_{\overline{t}}^{2^i} \mathrm{~(mod}~\psi_l) \mathrm{~~~~for~}i=0,1,2,\dots, \log_2 q,$$ each of
which will be a polynomial of degree less than ${l^2-1\over 2} = $ deg $~\psi_l$.  We then can
compute $x_{\overline{t}}^{q} \equiv x_{\overline{t}}^{2^{i_1}}\cdots x_{\overline{t}}^{2^{i_k}}$
(mod~$\psi_l$) where $q=2^{i_1}+\dots+2^{i_k}$ is the binary expansion of $q$.

Once we find $\overline{t}$ such that (\ref{Eqtosol}) is satisfied, we have found $\overline{t}$ such that
$$(x^{q^2},y^{q^2}) \oplus \overline{q}(x,y) = \pm \overline{t}(x^q,y^q) \mathrm{~(mod~}l).$$
In particular $-(x^q,y^q) = (x^q,-y^q)$ thus the $x$-coordinate can
only narrow down the possibilities of $t_l$ to two.

To find whether $t_l$ is $+\overline{t}$ or $-\overline{t}$ we check
whether or not
\begin{eqnarray} (y^\prime - y^q_{\overline{t}})/y \equiv 0 ~(\mathrm{mod~}\psi_l).\end{eqnarray}
If so, then we choose $+\overline{t}$, otherwise $t_l =
-\overline{t}$. This is sufficient for exactly the same reason as
the sufficiency of (\ref{Eqtosol}).

\subsection{The Cases  $(x^{q^2},y^{q^2})  = \pm\overline{q}(x,y)$}
\label{DecCases}

Our first important assertion is that if we falsely assumed points $(x^{q^2},y^{q^2})$ and 
$\overline{q}(x,y) = (x_{\overline{q}},y_{\overline{q}})$ had different 
$x$-coordinates, then the above procedure would not have found $\overline{t}$ such that (\ref{Ttosol}) was satisfied.  

Suppose on the contrary, that $\overline{t} \in \{1,2,\dots, {l-1\over 2}\}$ has satisfied (\ref{Eqtosol}) but that we have 
$(x^{q^2},y^{q^2})  = \pm\overline{q}(x,y)$.  Then 
$$\overline{t}(x^q,y^q)\ominus (x^{q^2},y^{q^2})$$ would be the same point as $\overline{q}(x,y) = \pm (x^{q^2},y^{q^2})$, modulo $l$.  But, this will 
lead to an immediate contradiction in the addition law.

So once we have failed to find a $\overline{t}$ we know that we have  $(x^{q^2},y^{q^2})  = \overline{q}(x,y)$ or 
$(x^{q^2},y^{q^2})  = -\overline{q}(x,y)$.  So now assume we are in this case.
Let us assume further that we have \begin{eqnarray} \label{secassump} (x^{q^2},y^{q^2})  = +\overline{q}(x,y).\end{eqnarray}  
Like before, our guess might be incorrect, but we will soon find out so.  
For $P = (x,y) \in \overline{E}[l]\setminus\{P_\infty\}$ satisfying (\ref{secassump}), 
we obtain via the characteristic equation (\ref{eqnchar}) modulo $l$ that 
$$\overline{t}\pi(P) = 2\overline{q}P.$$ By (\ref{secassump}), we also have $$\overline{t^2}\overline{q}P=\overline{t^2}\pi^2(P) = 
\overline{t}\pi(\overline{t}\pi(P))$$ which equals 
$(2\overline{q})^2P$.  Consequently, $$\overline{t}^2\overline{q}P \equiv (2q)^2P ~(\mathrm{mod~}l)$$ and thus $\overline{q}$ is a square modulo $l$ 
unless $\overline{t} \equiv 0 $ (mod $l$).  However, that would imply $2\overline{q}P = P_\infty$, which is a contridiction since $P \in \overline{E}[l]$ 
and gcd$(q,l)$ assumed to be $1$.

Once we know $q$ is a square, the rest follows easily.  We find $q$'s square-roots over $\f_l$ efficiently by using gcd$(x^2-q,x^l-x)$ or otherwise.
If $q \equiv w^2$ (mod $l$), we find $t_l$
will be $\pm 2w$ mod $l$ depending on the $y$-coordinate, i.e. the
rational function $y_w$.  \begin{eqnarray} \label{factchar} (X\pm w)(X\pm w) = X^2 \pm 2wX + q.\end{eqnarray} If $y_w$ matches $y^q$, then $t_l = 2w$,
otherwise $t_l = -2w$.  

If $q$ happened not to be a square, then our second assumption was also false, meaning we were in the third case 
$$(x^{q^2},y^{q^2})  = -\overline{q}(x,y).$$  Once we know we are in this csae, it is also easy since we immediately find 
$$(x^{q^2},y^{q^2})  \oplus \overline{q}(x,y) = P_\infty$$ since they were additive inverses.  Hence, $t_l \equiv 0$ in this case.

However, we still have to worry about the case of a false positive, i.e. it is possible for $q$ to be a square mod $l$ but 
$(x^{q^2},y^{q^2})  = -\overline{q}(x,y)$ nonetheless.  It is sufficient to check whether or not either square root $\pm w$ satisfies 
$\pi(P) = \pm wP$, for $P \in  \overline{E}[l]\setminus\{P_\infty\}$, hence 
leading to the factorization of the characteristic equation for $\pi$ as in (\ref{factchar}).   

The simple calculation of $$\gcd(\mathrm{numerator}(x^q-x_w),\phi_l)$$ and the test of whether or not it is $1$ suffices.
We have $\pi(P) = \pm wP$ for some $P \in  \overline{E}[l]\setminus\{P_\infty\}$ if and only if the gcd $\not = 1$.

\subsection{Case of l=2} \label{Case22}

Note that the above description actually works only for odd $l \not
= p$. However, we can also run an analogous procedure for $l=2$.
This allows us to choose set $S$ to contain slightly smaller primes.
Since we assume $q$ odd, $$q+1 - t \equiv t (\mathrm{mod~}2)$$ and
in particular, $t_2 \equiv 0 ~($mod $2)$ if and only if $E(\f_q)$
has an element of order $2$.  By definition of $\oplus$, any element
of order $2$ must be of the form $(x_0,0)$.  Thus $t_2 \equiv 0
~($mod $2)$ if and only if $x^3+Ax+B$ has a root in $\f_q$.  An
efficient way to check this is by taking the gcd$(x^q-x, x^3+Ax+B)$.
Here we compute $x^q$ efficiently by taking successive squares
modulo $x^3+Ax+B$.  In summary $t_2 \equiv 0$ if and only if
gcd$(x^q-x, x^3+Ax+B) \not = 1.$

\subsection{Summary}
We can summarize the procedure as follows:

1) Choose a set of primes $S$, with $p \not\in S$ such that $\prod_{l \in S} l > 4\sqrt{q}$.

2) For $l = 2$, we find $t_l \equiv 0$ if and only if gcd$(x^q-x, x^3+Ax+B) \not = 1.$

3) For $l \in S\setminus\{2\}$, do the following:

\hspace{2em} a) Let $\overline{q}$ be the unique integer satisfying $\overline{q} \equiv q$ (mod $l$) and $|\overline{q}|< {l \over 2}$.

\hspace{2em} b) Compute univariate rational function $x^\prime$ as defined above.  We can even work  modulo $\psi_l$ so we can use polynomials of 
bounded degree instead.

\hspace{2em} c) For $\overline{t} \in \{1,2,\dots, {l-1\over 2}\}$ do:

\hspace{4em}   ~i) Check if $x^\prime - x_{\overline{t}}^q \equiv 0$ modulo $\psi_l$.  If so, go to step (ii).  Otherwise, try the next value of 
$\overline{t}$.  If you have unsuccessfully tried ${l-1\over 2}$, go to step (d) instead. 

\hspace{4em}  ii) Check whether or not $(y^\prime-y_{\overline{t}}^q)/y \equiv 0$ modulo $\psi_l$.  If so, then $t_l \equiv \overline{t}$.  Otherwise, 
$t_l \equiv -\overline{t}$. 

\hspace{2em} d) Find $q$'s square roots modulo $l$, if they exist.  If they don't exist then $t_l \equiv 0$.  Otherwise write $q \equiv w^2$ (mod $l$).  

\hspace{2em} e) Check whether or not gcd(numerator($x^q-x_w),\psi_l)=1$.  If so, $t_l \equiv 0$.  Otherwise go to step (f).

\hspace{2em} f) Check whether or not gcd(numerator($(y^q-y_w)/y),\psi_l)=1$.  If this gcd is $1$, then $t_l \equiv -2w$.  Otherwise, $t_l \equiv 2w$.

4) We now have computed $t_l$ for all $l \in S$.  Thus we know $|E(\f_q)| \equiv 1 + q - t_l ~($mod $l$) for every $l$ in $S$.  
Using the Chinese Remainder 
Theorem, we obtain $|E(\f_q)|$ modulo $N$, where $N = \prod_{l\in S} l$.  
Since set $S$ was chosen so that $N > 4\sqrt{q}$, by Hasse's theorem, we in fact know 
$|E(\f_q)|$ precisely.
\vspace{1em}

As described in the introduction, Schoof's algorithm was a huge
improvement over previous methods, having an asymptotic running
times of $(\log q)^8$.  In the 1990's Noam Elkies discovered an
improvement to this algorithm that yields a faster running time. The
trick is to restrict set $S$ to primes of a certain type, now known
as Elkies primes, and to use modular polynomials instead of division
polynomials.  Before describing this improvement, we will first
introduce modular polynomials.

\section{Modular Polynomials}

Modular Polynomials come from the theory of modular forms and an
interpretation of elliptic curves over $\C$ as lattices.  Being a
lattice, the matrix group $SL_2(\z)$ naturally acts on $\C/(\z +
\z\tau)$.  Two lattices are considered to be equivalent if there is a natural way to re-scale
and rotate one to get the other. It turns out that two elliptic
curves are isomorphic as groups if their associated $\C$-lattices
are equivalent.  Thus a certain invariant, known as the {\bf
$j$-invariant}, of lattice theory is in fact an invariant of
isomorphic elliptic curves as well.  For $E$ given in
Weierstra\ss~form (\ref{WeEq}), we have $$j(E) =1728{4A^3 \over 4A^3
+ 27B^2}.$$

\subsection{Thinking in terms of cosets}

We will now go back and forth between $j(E)$ and $j(z)$, where we
use $j(z)$ to refer to the $j$-invariant of a lattice with basis
$\{\omega_1,\omega_2\}$ such that $z = {\omega_1 \over \omega_2} \in
\C$.  The matrix groups $GL_2(\z)$ and $SL_2(\z)$ naturally act on
$\C$ so it makes sense to use notation such as $j(M\cdot z)$ for $M
\in GL_2(\z)$ or $SL_2(\z)$.

\begin{Prop} \cite{Stark}
If $M$ is a $2\times 2$ integer matrix with $\det M = m  \in
\z_{>0}$ then $j(M\cdot z)$ and $j(z)$ are algebraically related,
meaning there exists a bivariate polynomial $\Phi_m(x,y) \in
\z[x,y]$ such that $\Phi_m\bigg(j(M\cdot z), j(z)\bigg) = 0$.
\end{Prop}

\begin{proof} (Sketch)


%Supposing that $M$ is in upper-triangular form, which implies $$M =\bigg(\begin{matrix}a~b \\ 0~d
%\end{matrix}\bigg) \mathrm{~~where~}ad = m.$$

If we consider the group $M_m$ of all matrices of determinant $m$,
we obtain that $M_m$ has a left-coset decomposition as
$$M_m = \bigcup_{a,b,d} SL_2(\z)\cdot \bigg(\begin{matrix}a~b \\ 0~d\end{matrix}\bigg)$$ where we add the
restrictions $a > 0,~ d > 0,~ ad = m,$ and $0 \leq b < d$.

Since the set of $j$-invariants $$\bigg\{j(B\cdot z) ~ :~ B \in
M_m\bigg\} = \bigg\{j(BA\cdot z) ~ :~ B \in M_m\bigg\}
\mathrm{~for~}A\in SL_2(\z),$$ we find that
$$\Phi_m(x,j(z)) = \prod_{M\in M_m} \bigg(x- j(M\cdot z)\bigg) = \bigg(x - j\bigg({az + c \over d}\bigg)\bigg)$$
satisfies the desired properties.  Note that it can be shown that this polynomial is integral.
\end{proof}

%\begin{Prop} The cardinality of $M_m$ is $n\prod_{p|n} (1+{1\over p})$ hence we obtain the degree
%of $\Phi_m$ equals  $$2\sum_{\stackrel{c|m}{c > \sqrt{m}}} c + \delta_{\sqrt{m}}\cdot(\sqrt{m}-1)$$
%where $\delta_{\sqrt{m}} = 1$ if $m$ is a perfect square and equals $0$ otherwise.
%\end{Prop}

%\begin{proof}
%FROM \cite[Nov. 7]{Stark}
%\end{proof}

%HERE, I SHOW/SUMMARIZE THAT it makes sense to define a group

%$$S_n^* = \bigg\{\bigg(
%\begin{matrix}a~b \\ 0~d
%\end{matrix}
%\bigg) \in D_n^* : d > 0,~0 \leq b < d\bigg\}$$ with cardinality $n\prod_{p|n} (1+{1\over p})$
%and polynomial

%$$\Phi_n(x,j) = \prod_{\alpha \in S_n^*} (x-j\cdot \alpha),$$ a polynomial in $\z[x,j]$ which is symmetric in each variable, and degree
%$\# S_n^*$ in both variables.  This is known as a/the {\bf modular polynomial}.  My main sources are \cite{Blake} and \cite{Stark}.%

%The polynomials $\Phi_n(x,j)$ have the remarkable property, known as the {\bf Kronecker congruence relation} that
%$$\Phi_l(x,y) \equiv (x^l-y)(x-y^l) \mathrm{~(mod~}l).$$

%The degrees of modular polynomials are only $l+1$ rather the degree of $(l^2-1)/2$ of the divison polynomials.

\section{Sketch of Elkies' Improvement}

Given the characteristic equation for the Frobenius map, $$\pi^2 -
t\pi + q = 0,$$ we define a prime $l$ to be an {\bf Elkies prime} if
this equation splits over $\f_l$.  This is equivalent to whether or
not the discriminant $t^2-4q$ is a square modulo $l$.  We define $l$
to be an {\bf Atkin prime} if it is not an Elkies prime.

\begin{Rem} Note that the condition of being an Elkies or Atkin prime is
contingent on the choice of curve $E$ and field $\f_q$ which
determine a specific characteristic equation for $\pi$.  Since we
have assumed a fixed choice of $E$ and $\f_q$ all along, we will
from here on out simply refer to primes $l$ as Elkies or Atkin
primes, without further specification, for the sake of expository
convenience.
\end{Rem}

Since our goal is to compute $t$, which is required to test what
type a given prime is, this definition has yet to improve the
algorithm; however this is where the modular polynomial $\Phi_l$ for
prime $l$ plays a role.  The following deep result plays a key role.
We omit its proof but please see \cite{Blake} or \cite{Dewaghe} for
details.

\begin{Prop}
A prime $l$ is an Elkies prime for curve $E$ over $\f_q$ if and only if $\Phi_l(x,j(E))$ has a root
for $x \in \f_q$, where $j(E)$ is the $j$-invariant for curve $E$.
\end{Prop}

It is efficient to test whether or not prime $l$ is an Elkies prime
or not: we take the gcd of $\Phi_l(x,j(E))$ with $x^q-x$ as in
Section \ref{Case22}.  If $l$ is not an Elkies prime, we do not
include it in our set $S$. Furthermore, roughly half the primes are
Elkies primes (a non-trivial but known fact), so this will not
increase the size of the primes in $S$ too significantly.  (In all
this, we always assume that $l \not = 2, p$ where $p$ is the
characteristic of $\f_q$.)  The added efficiency for computing $t$
(mod $l$) for only Elkies primes $l$ will dominate the reduced
efficiency for using larger primes $l$.

In particular, when $l$ is an Elkies prime, we will be able to solve
the equation $\pi(x,y) = (x^q,y^q) = \lambda(x,y)$ for $\lambda$
using a degree ${l-1 \over 2}$ factor of the degree ${l^2 - 1 \over
2}$ division polynomial $\psi_l$.  By virtue of being an Elkies
prime, the characteristic equation for $\pi$ splits modulo $l$, and
it follows that
$$\pi^2 - t\pi + q = (\pi-\lambda t)(\pi - \overline{\lambda}t)$$
such that $\lambda\overline{\lambda}=q$.  Consequently, $$t_l \equiv
\lambda + {q \over \lambda} ~(\mathrm{mod~}l).$$

Thus to compute $t_l$, it is sufficient to compute an eigenvalue
$\lambda \in \{1,2,\dots,l-1\}$ of the Frobenius map, and hence that
is now our current goal.
%When $l$ is an Elkies prime, this can be accomplished by using a
%degree $(l-1)/2$ of the division polynomial $\psi_l$.
For this purpose, we consider a cyclic subgroup of $E(\f_q)$, which
we denote by $C$ which is fixed by the action of the Frobenius map.
Thus the polynomial
$$F_l(x) = \prod_{\pm P_i \in C \setminus \{P_\infty\}} (x- (P_i)_X)$$ has integral coefficients, hence
is defined over $\f_q$.  Here this product is meant to only include
one out of each of the pairs $\pm P_i$, since both have the same
$x$-coordinate.  The notation $(P_i)_X$ signifies the $x$-coordinate
of point $P_i$.  Note that there are efficient methods for
explicitly computing $F_l(x)$, we refer the reader to Section VIII.4
of \cite{Blake}.

We now can work modulo $F_l(x)$ instead of modulo $\psi_l$ which has
smaller degree, ${l-1 \over 2}$, hence is more efficient; i.e.
instead of needing to solve equation (\ref{Eqtosol}), it suffices to
solve
\begin{eqnarray} x^\prime - x^q_{\lambda} \equiv 0 ~(\mathrm{mod~}F_l)\end{eqnarray} for
$\lambda$.
%To compute $t$ (mod $l$) for an Elkies prime, we utilize the fact that if $x=\alpha \in \f_q$ is a
%root of $\Phi_l(x,j(E))$, then it follows that $\alpha$

We end with the note that Atkin's improvement involves the efficient
extraction of information about $t_l$ even for $l$ a non-Elkies
prime, i.e. an Atkin prime.  However the description of this
algorithm is a topic for another day.  Please see \cite{Blake} or
\cite{Schoof2} for details.



%FILL IN HOW WE COMPUTE $t$ (mod $l$) FOR AN ELKIES PRIME!!!
%
%\vspace{30em}
%For roughly half the primes, property FILL IN holds.  Such a prime
%is now known in the literature as an Elkies prime.

%For such a prime, we can work with modular polynomials instead of division polynomials to solve equations analogous to (\ref{Eqtosol}) for
%$\overline{t}$.  Since modular polynomials have roughly half the degree of division polynomials, this leads to a speed-up of the algorithm.

%MORE DETAIL NEEDED.

%Also, it is not a problem that all primes are not Elkies primes
%since we just choose set $S$ accordingly.


%EXAMPLE




\begin{thebibliography}{99}

\bibitem{Blake} I. Blake, G. Seroussi, and N. Smart.  \emph{Elliptic Curves in Crytography}, volume 265 of 
\emph{London Mathematical Society Lecture Note Series}.  Cambridge University Press, Cambridge, 2000.  Reprint of the 1999 original.

\bibitem{Dewaghe} L. Dewaghe.  Remarks on the Schoof-Elkies-Atkin Algorithm. \emph{Math. Comp.,} 
{\bf 67}(223):1247-1252, 1998.

\bibitem{CRT}  D. Dummit and R. Foote.  \emph{Abstract Algebra, 2nd Edition.}  Prentice Hall, Upper Saddle River, 1999.

\bibitem{Enge} A. Enge. \emph{Elliptic Curves and their Applications to Cryptography: An Introduction.}  
Kluwer Academic Publishers, Dordrecht, 1999.

\bibitem{Harris} N. Harris.  \emph{Math 168 Project.} University of California, San Diego, Fall 2005.

\bibitem{Lang} S. Lang. \emph{Elliptic Curves: Diophantine Analysis.} Springer-Verlag, Berlin, 1978.

\bibitem{PariImp}
http://pari.math.u-bordeaux.fr/archives/pari-announce-05/msg00002.html

\bibitem{Schoof1} R. Schoof.  Elliptic Curves over Finite Fields and the Computation of Square Roots mod $p$. \emph{Math. Comp.,} 
{\bf 44}(170):483-494, 1985.

\bibitem{Schoof2} R. Schoof.  Counting Points on Elliptic Curves over Finite Fields. \emph{J. Th\`eor. Nombres Bordeaux} {\bf 7}:219-254, 1995.

\bibitem{Stark} H. Stark. \emph{Lecture notes for Math 204.} University of California, San Diego, Fall 2005.

\bibitem{Wash} L. C. Washington, \emph{Elliptic Curves: Number Theory and Cryptography}.  Chapman \& Hall/CRC, New York, 2003.

\end{thebibliography}

\end{document}
